\appendix

\section{DPGrammar Algorithm (\texttt{dp\_fix\_prefix})}
\label{app:alg}

Algorithm~\ref{alg:dpgrammar} gives the full pseudocode for the Viterbi segment repair used by DPGrammar.
Given a violated span $[c,s)$ detected during generation, the algorithm maintains a dictionary of live DFA states, each mapped to its highest-scoring path (matcher clone and cumulative log-probability).
At each position it probes the top-$K$ vocabulary entries via rollback—no cloning—and records the best (prev\_state, token) pair for each reachable DFA state (Viterbi merge).
After the loop, it clones one matcher per surviving state (Phase~2) and records the backtrack pointers.
Backtracking from the best final state yields the minimal set of token replacements needed to make the span grammar-valid.

Two design choices keep the algorithm efficient.
First, DFA states are identified by \texttt{bytes(matcher.compute\_logit\_bias())}, a compact proxy for the automaton node; this bounds the number of active states by the grammar's DFA size ($\sim$20--200 for typical JSON schemas) rather than by $K^{\text{depth}}$.
Second, matcher clones are created only once per surviving state per step (Phase~2), not once per (state, candidate) pair; cloning cost is therefore $O(|\text{winners}|) \leq O(\text{DFA size})$ per position rather than $O(K \times |\text{states}|)$.

The span $[c,s)$ is bounded by \texttt{find\_constraint\_end}, which probes forward from $c$ using the grammar automaton and tracks bracket depth in the original token sequence to avoid stopping inside an unclosed array or object.
The DP is thus applied to a narrow window around the violated constraint rather than the full remaining sequence.

\IfFileExists{algpseudocode.sty}{
\begin{algorithm*}[t]
\caption{DPGrammar: Viterbi segment repair over a violated span (\texttt{dp\_fix\_prefix})}
\label{alg:dpgrammar}
\begin{algorithmic}[1]
\Require Violated span $[c,s)$; log-probabilities $\ell \in \mathbb{R}^{L \times |V|}$ from the current forward; grammar matcher $\mathcal{M}$ at the DFA state just before $c$; beam width $K$
\Ensure Replacement pairs $\{(p_i, t_i)\}$ maximising $\sum \log p(t_i)$ subject to grammar, or \textsc{Fail}
\State $P \gets [c,\, c{+}1,\, \ldots,\, s{-}1]$
\State $\mathit{states} \gets \{\operatorname{key}(\mathcal{M}) \mapsto (\mathcal{M},\ 0)\}$ \Comment{DFA-state $\to$ (matcher, cumulative log-prob)}
\State $\mathit{back} \gets \{\}$ \Comment{backtrack table}
\For{$i \gets 0$ \textbf{to} $|P|-1$}
    \State $T_K \gets \operatorname{top\text{-}K}(\ell[P[i]],\; k{=}K)$
    \State $\mathit{winners} \gets \{\}$
    \For{\textbf{each} $(\text{prev\_key},\, (\mathcal{M}_p,\, \sigma_p))$ \textbf{in} $\mathit{states}$}
        \For{\textbf{each} $(t,\, \ell_t)$ \textbf{in} $T_K$}
            \If{$\mathcal{M}_p.\operatorname{try\_consume}([t]) < 1$}
                \State $\mathcal{M}_p.\operatorname{rollback}(0)$; \textbf{continue}
            \EndIf
            \State $\sigma' \gets \sigma_p + \ell_t$;\quad $k' \gets \operatorname{key}(\mathcal{M}_p)$
            \State $\mathcal{M}_p.\operatorname{rollback}(1)$
            \If{$k' \notin \mathit{winners}$ \textbf{or} $\mathit{winners}[k'].\sigma < \sigma'$}
                \State $\mathit{winners}[k'] \gets (\text{prev\_key},\; t,\; \sigma')$ \Comment{Viterbi merge}
            \EndIf
        \EndFor
    \EndFor
    \If{$\mathit{winners} = \emptyset$}\ \Return \textsc{Fail} \EndIf
    \State $\mathit{next} \gets \{\}$
    \For{\textbf{each} $(k',\, (\text{prev\_key},\, t,\, \sigma'))$ \textbf{in} $\mathit{winners}$} \Comment{Phase 2: clone once per surviving state}
        \State $\mathcal{M}' \gets \mathit{states}[\text{prev\_key}].\mathcal{M}.\operatorname{deep\_copy}()$;\quad $\mathcal{M}'.\operatorname{try\_consume}([t])$
        \State $\mathit{next}[k'] \gets (\mathcal{M}',\; \sigma')$;\quad $\mathit{back}[(i,\,k')] \gets (\text{prev\_key},\; t)$
    \EndFor
    \State $\mathit{states} \gets \mathit{next}$
\EndFor
\State $k^* \gets \arg\max_k\; \mathit{states}[k].\sigma$;\quad $\mathit{fixes} \gets []$;\quad $\text{cur} \gets k^*$
\For{$i \gets |P|-1$ \textbf{downto} $0$}
    \State $(\text{prev},\; t) \gets \mathit{back}[(i,\;\text{cur})]$
    \If{$t \neq x[P[i]]$}\ $\mathit{fixes}.\operatorname{append}(P[i],\; t)$ \EndIf
    \State $\text{cur} \gets \text{prev}$
\EndFor
\State \Return $\operatorname{reverse}(\mathit{fixes})$
\end{algorithmic}
\end{algorithm*}
}{}

\section{Case Study: Instance \texttt{o12610}}
\label{app:case}

We examine instance \texttt{o12610} from \texttt{jsb\_medium}, a representative case where LAVE completely fails while DPGrammar produces a clean, fully valid output in a single pass with zero resamples.

\paragraph{Schema.}
The schema describes a physical Wi-Fi radio device configuration (OpenWRT/LuCI format).
The critical constraints are enum fields on \texttt{htmode} and \texttt{hwmode}:

\begin{lstlisting}[style=appendixcode]
"htmode": { "type": "string",
  "enum": [null, "HT20", "HT40+", "HT40-",
           "VHT20", "VHT40", "VHT80"] },
"hwmode": { "type": "string",
  "enum": ["11b", "11g", "11a"] }
\end{lstlisting}

Additional fields include \texttt{channel} (\texttt{integer}, range 1--165), \texttt{id} (string), \texttt{short\_gi\_20/40/80} (boolean), and \texttt{rx\_stbc}/\texttt{tx\_stbc} (integer, 0--1).

\paragraph{Why LAVE fails (timeout, 120\,s, no output).}
LAVE's stochastic lookahead commits a token only if all $K$ randomly sampled completions yield a grammar-valid continuation.
For enum fields, this requires a random completion to land on exactly one of a small set of technical strings (e.g.\ \texttt{"VHT20"}, \texttt{"11a"}) drawn from a natural-language vocabulary.
The probability is near zero: LLaDA-8B assigns negligible probability mass to \texttt{"VHT20"} or \texttt{"HT40-"} as free continuations, so every lookahead batch fails the grammar check.
LAVE cannot commit even the first enum-constrained token and times out after 120\,s without producing any output.

\paragraph{DPGrammar (valid, 0 resamples, 6.7\,s).}
Frontier masking restricts the vocabulary at each position to the set of tokens permitted by the grammar automaton.
For the \texttt{htmode} field, only the four tokens spelling out a valid enum value are unmasked; the model samples from this constrained set directly.
Because the automaton guides every token choice, no violation occurs and the Viterbi DP is never triggered (0 resample events, 0 DP calls).
The output is complete and semantically coherent:

\begin{lstlisting}[style=appendixcode]
{
    "channel": 60,
    "htmode": "VHT20",
    "hwmode": "11a",
    "id": "pci-wifi-1",
    "rx_stbc": 1,
    "short_gi_20": true,
    "short_gi_40": true,
    "short_gi_80": false,
    "tx_stbc": 1
}
\end{lstlisting}

All enum constraints are satisfied (\texttt{htmode} $\in$ allowed set, \texttt{hwmode} $=$ \texttt{"11a"}), the channel is within range, and boolean/integer fields are correctly typed.
Total wall time is 6.7\,s versus LAVE's 120\,s timeout—an 18$\times$ speedup on this instance.

\paragraph{Summary.}
This instance illustrates the general failure mode of stochastic lookahead on enum-constrained fields: the probability of a random natural-language completion matching a specific technical string (e.g.\ \texttt{"VHT20"}) is effectively zero, so LAVE never commits.
Frontier masking (shared by Dgrammar and DPGrammar) sidesteps this entirely by restricting the model's sampling distribution to the valid set at each position.
On instances where frontier masking prevents all violations, DPGrammar's Viterbi repair is never invoked—the method degrades gracefully to pure frontier-masked generation at the speed of a single unconstrained diffusion pass.

\section{Hyperparameters}
\label{app:hyperparams}

All experiments use LLaDA-8B-Instruct as the backbone, run on Modal with NVIDIA A100 GPUs.
Table~\ref{tab:hyperparams} lists the shared settings across all three methods.

\begin{table}[h]
\centering
\small
\begin{tabular}{ll}
\toprule
\textbf{Hyperparameter} & \textbf{Value} \\
\midrule
Diffusion steps $T$          & 128 \\
Random seed                  & 0 \\
Block length                 & 32 \\
Remasking strategy           & Low-confidence \\
Temperature                  & 0.2 \\
Top-$K$ (DPGrammar DP)       & 100 \\
Max resample attempts        & 500 (LAVE), 200 (ours) \\
Timeout threshold            & 120\,s \\
\bottomrule
\end{tabular}
\caption{Shared hyperparameters for all benchmark runs on \texttt{jsb\_medium}.}
\label{tab:hyperparams}
\end{table}

\section{Dataset and Instance Filtering}
\label{app:data}

We evaluate on the \texttt{Github\_medium} split of JSONSchemaBench, which contains 586 instances.
Of these, 75 are excluded from all comparisons because \textsc{llguidance} rejects their schemas at compile time: LAVE immediately returns \texttt{valid=false} with the error \texttt{"Prompt does not conform to grammar"}; Dgrammar and DPGrammar silently skip them via the same \textsc{llguidance} guard.
Since all three methods behave identically on these 75 instances, including them would inflate LAVE's failure count without providing a meaningful comparison.
All results in Table~\ref{tab:main} are therefore reported on the matched $n{=}511$ subset.

\section{Timing Breakdown}
\label{app:timing}

Table~\ref{tab:timing} reports per-method timing statistics supplementing Table~\ref{tab:main}.
Forward time dominates in all cases; constraint overhead is a small fraction for Dgrammar and DPGrammar.
LAVE's higher mean time is driven by timeout instances (capped at 120\,s); its median is lower because the majority of instances either succeed quickly or exhaust their resample budget early.
Effective constraint time for DPGrammar is near zero on instances where no DP repair is triggered; the non-trivial mean reflects the small fraction of instances that require one or more Viterbi passes.

\begin{table*}[h]
\centering
\small
\begin{tabular}{lrrr}
\toprule
\textbf{Metric} & \textbf{LAVE} & \textbf{Dgrammar} & \textbf{DPGrammar} \\
\midrule
Avg fwd passes              & 99.2  & 104.7 & 99.4  \\
Constraint overhead (\%)    & 11.95 & 24.25 & 12.67 \\
Eff.\ constraint (\%)       & —     & 20.43 & 1.22  \\
Per-token total (ms)        & 111.5 & 100.7 & 71.9  \\
Per-token constraint (ms)   & 6.64  & 8.26  & 0.30  \\
Avg mask compute (ms)       & —     & 427.2 & 577.0 \\
Avg forward total (ms)      & 3951  & 4612  & 4229  \\
\bottomrule
\end{tabular}
\caption{Detailed timing breakdown on \texttt{jsb\_medium} ($n{=}511$ matched instances).
Effective constraint \% measures time spent in grammar-check calls that were on the critical path; it is not available for LAVE, which lacks the same instrumentation.}
\label{tab:timing}
\end{table*}
